---
title: "Edulevel"
author: "Xiangzhen Li, Xiangmei Zhang"
date: "2017/4/25"
output: 
  rmdshower::shower_presentation:
    self_contained: false
    katex: true
    theme: ribbon
---
```{r setup, include=FALSE, message=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```
##Edulevel:Temporal changes and spatial distribution of educational level for the US

Xiangmei Zhang & Xiangzhen Li

##Edulevel

- Data clean
- Package
- Shiny app

##Data clean

- Data source: Four educational level of all US counties from 1970-2015;
- Data: 3283 observation * 47 variables;
- source: <https://www.data.gov/>.


```{r warning = FALSE, message= FALSE}
dat <- readxl::read_excel("../Education.xls", skip = 4)
colnames(dat)
```



## Data clean

- Tidy format
- Consistency check
- Correctness check

```{r}
library(Edulevel)
head(tidy)

```

##Package: `Edulevel` 

- Visualization tools to analysis the temporal changes and spatial distribution of educational level for the US
- Four functions: *lineEd*, *showcounty*, *mapstate*, *mapcounty*. 
- One dataset: *tidy*.

##Package: `Edulevel`
*showcounty* : Show list of counties of a given state. County names will be used in other functions.
```{r warning = FALSE, message= FALSE}
library(ggplot2)
library(tidyverse)
library(rvest)
library(plotly)

showcounty("IA")
```

##Package: `Edulevel`
*lineEd* : Plot the education level of a given county.
```{r warning = FALSE, message= FALSE, fig.height=4}
#Story county in Iowa
lineEd("IA", "Story County")
```

##Package: `Edulevel`
The education levels is different from county to county.
```{r warning = FALSE, message= FALSE,fig.height=4}
#Liberty county in Florida
ggplotly(lineEd("FL", "Liberty County"))
```

##Package: `Edulevel`
*mapstate* :Map the education level of all states.
```{r warning = FALSE, message= FALSE,fig.height=4.5}
#Education level: 4="Bachelor's degree or higher".
mapstate(vtype = "percent", levelint = 4)
```

##Package: `Edulevel`
```{r warning = FALSE, message= FALSE}
#Education level: 1="Less than a high school diploma"
ggplotly(mapstate(vtype = "percent", levelint = 1))
```

##Package: `Edulevel`
"percent" is better than "count" in here.
```{r warning = FALSE, message= FALSE,fig.height=4.5}
#Education level: 1="Less than a high school diploma"
ggplotly(mapstate(vtype = "count", levelint = 1))
```

##Package: `Edulevel`
*mapcounty* Map the education level of a given state.
```{r warning = FALSE, message= FALSE,fig.height=4.5}
#limit: Label county names whose percentage > limit.
mapcounty(vtype = "percent", levelint = 4,stateshort="IA", limit=30)
```

##Package: `Edulevel`
*mapcounty* Map the education level of a given state.
```{r warning = FALSE, message= FALSE,fig.height=4.5}
#Change limit to 40
mapcounty(vtype = "percent", levelint = 4,stateshort="IA", limit=40)
```


##Shiny app

Shiny: Edulevel
```{r warning = FALSE, echo=T,eval=FALSE, message=FALSE}
shiny::runGitHub("585-project", "Xiangmei21", 
                 subdir = "Project/shiny.R")
```