---
title: "Edulevel"
author: "Xiangmei Zhang"
date: "3/31/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Edulevel package

This package contains one data set and 4 functions. data set "tidy" is the cleaned data of Education levels of adults more than 25 years old in US from 1970 to 2015. Function "showcounty" returns a list of counties for a given state abbrevation, countyname can be used in "lineEd" function, "lineEd" plots all education level lines for a given county, "mapstate" function maps the "count" or "precent" data on the scale of states in US for a given education level, "mapcounty" function maps maps the "count" or "precent" data on the scale of counties for a given state and a given education. level

```{r cars}
library(Edulevel)
library(tidyverse)
state.abb
```
```{r}
showcounty("AL")
```
```{r}
lineEd("AL","Perry County")
```
```{r}
mapstate(vtype="percent",levelint = 4)
```
```{r}
mapcounty(levelint = 4,stateshort = "AL",limit = 30)
```
